---
title: "Assignment_5_Stats_II_Group_22"
author:
- "Maik David Klaiber (22-482-426)"
- "Jonas Weber (24-621-484)"
- "Yann Rochaix (24-622-276)"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
output: html_document
---

# Data cleaning procedure

The same data cleaning procedure was used as in Exercise 5.

```{r setup, include=FALSE}

# Load required packages
library(skimr)
library(dplyr)
library(stargazer)
library(lfe)
library(marginaleffects)

```

```{r read csv, include=TRUE}
# Load the data set
df <- read.csv("education_income-3.csv")

```

```{r data cleaning, include=TRUE, results='hide'}
# Summarize the data
skim(df)

# Cleaning

df <- df %>%
  
# Replace invalid values with NA
  
  mutate(
    college = ifelse(college == -1, NA, college),
    
    income = ifelse(income == -9999999, NA, income),
    hhincome = ifelse(hhincome == -9999999, NA, hhincome),
    
    # Create new variables
    white = race,
    
    low_income = ifelse(
      hhincome < quantile(hhincome, 0.5, na.rm = TRUE),
      1,
      0
    )
  ) %>%
  
# Remove old variable
  select(-race)

# Summarize cleaned data
skim(df)
```

# Question 1: Descriptive Analysis

```{r descriptive statistics, include=TRUE}


income_college <- 1000 * mean(df[df$college==1,'income'], na.rm = TRUE)
income_noCollege <- 1000 * mean(df[df$college==0,'income'], na.rm = TRUE)
diff <- income_college - income_noCollege

# Create summary table
summary_table <- data.frame(
  Group = c("College Degree", "No College Degree", "Difference"),
  Average_Income = c(income_college, income_noCollege, diff)
)

# Display table
knitr::kable(
  summary_table,
  digits = 0,
  col.names = c("Group", "Average Income (USD)"),
  format.args = list(big.mark = ",")
)


```

<b>(a) Can we interpret raw mean income difference between college and non-college individuals as the causal effect of attending college on income? Why or why not?</b>

No, we cannot interpret the raw mean income difference of `r format(round(diff,0), big.mark=",")` USD as the causal effect of attending college on income. Individuals who attend college may systematically differ from non-attendees in observable and unobservable characteristics such as ability, motivation, or family background.

These factors affect both college attendance and income, violating the exogeneity assumption $E(u∣college)=0$. Hence, the observed difference reflects both the causal effect of college and selection bias due to omitted variables. $$\text{Observed Difference} = \text{Causal Effect} + \text{Selection Bias}$$.

# Question 2: Assignment Mechanism

<b>(a) Decomposition of the Observed Mean Difference into ATET and Selection Bias</b>

A researcher estimates the OLS regression

$$
income_i = \beta_0 + \beta_1 college_i + u_i
$$

and obtains $\hat{\beta}_1 > 0$.

We decompose the observed mean difference as follows:

$$
E(Y_i \mid D_i = 1) - E(Y_i \mid D_i = 0) = 
\underbrace{E(Y_{1i} - Y_{0i} \mid D_i = 1)}_{\text{ATET}} + 
\underbrace{\left[ E(Y_{0i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0) \right]}_{\text{Selection Bias}}
$$

<b>(b) Do you think that in this setting $\hat{\beta}_1$ measures the causal effect of interest or is it a biased estimate? Briefly explain your answer.</b>

No, $\hat{\beta}_1$ is likely a biased estimate rather than the causal effect of interest. For a causal interpretation, the exogeneity assumption $E(u | college)=0$
must hold. However, individuals who attend college may systematically differ from non-attendees in characteristics such as ability, motivation, or family background. Since these factors may affect both college attendance and income, the error term is correlated with the college variable, leading to omitted variable bias and selection bias.

<b>(c) If you think that it is a biased estimate, which variable in the dataset could be a key driver of selection into college? If you think the estimate is correct, skip the following questions.</b>

A potential key driver of selection into college is household income (`hhincome`). Individuals from higher-income households may have better access to education, financial support, and academic opportunities, making them more likely to attend college. At the same time, household income may also be related to future earnings, which can lead to omitted variable bias.

<b>(d) Regression with Observed Controls and Selection-on-Observables</b>

We first estimate the baseline regression:

$$
income_i = \beta_0 + \beta_1 college_i + u_i
$$
```{r baseline regression, include=TRUE}
model_1 <- felm(income ~ 1 + college, data = df)

summary(model_1, robust = TRUE)

```

We then estimate the controlled regression (control for`hhincome`, `race` and `year`:

$$
income_i = \beta_0 + \beta_1 college_i + \beta_2 hhincome_i + \beta_3 race_i + \beta_4 year_i + u_i
$$


```{r controlled regression, include=TRUE}
model_2 <- felm(
  income ~ 1 + college + hhincome + white + year,
  data = df
)

summary(model_2, robust = TRUE)
```

```{r regression_table, echo=FALSE}

# Baseline regression
model_1 <- felm(income ~ college, data = df)

# Regression with controls
model_2 <- felm(
  income ~ college + hhincome + white + year,
  data = df
)

# Create comparison table
comparison_table <- data.frame(
  Model = c("Baseline", "With Controls"),
  College_Coefficient = c(
    coef(model_1)["college"],
    coef(model_2)["college"]
  )
)

# Display table
knitr::kable(
  comparison_table,
  digits = 3,
  col.names = c("Model", "Coefficient on College")
)

```

<b>Does adding these controls address the selection-on-observables concern?</b>

After adding the controls, the estimated coefficient on college decreases from 41.499 in the baseline regression to 28.391 in the controlled regression. This suggests that part of the original income difference was driven by observable characteristics such as household income, race, and year. Therefore, adding these controls helps address selection-on-observables concerns. However, unobserved factors such as ability or motivation may still violate exogeneity, so the coefficient cannot yet be interpreted as fully causal.

<b>(e) Explain why random assignment of scholarship would eliminate the selection bias term from part (a).</b>

Random assignment eliminates selection bias because scholarship amounts are assigned independently of individual characteristics such as ability, motivation, or family background. Therefore, scholarship assignment is no longer correlated with omitted factors in the error term. In other words,

$$
Cov(\text{scholarship}, u)=0
$$

and the exogeneity assumption holds. As a result, the selection bias term from part (a) disappears, since treated and untreated individuals are comparable on average.

<b>(f) Estimate a two-stage least squares model using scholarship as an instrument for college, controlling for household income, race, and year.</b>

We estimate the following first-stage regression:

$$
\text{college}_i = \pi_0 + \pi_1 \text{scholarship}_i + \pi_2 \text{hhincome}_i + \pi_3 \text{white}_i + \pi_4 \text{year}_i + e_i
$$

```{r first_stage, include=TRUE}

first_stage <- felm(
  college ~ 1 + scholarship + hhincome + white + year,
  data = df
)

summary(first_stage, robust = TRUE)

```

We then estimate the two-stage least squares (2SLS) model:

$$
\text{income}_i = \beta_0 + \beta_1 \text{college}_i + \beta_2 \text{hhincome}_i + \beta_3 \text{white}_i + \beta_4 \text{year}_i + u_i
$$

using `scholarship` as an instrument for `college`.

```{r iv_regression, include=TRUE}

iv_model <- felm(
  income ~ 1 + hhincome + white + year |
  0 |
  (college ~ scholarship),
  data = df
)

summary(iv_model, robust = TRUE)

```

- **Comment on first stage regression:** The first-stage regression shows that scholarship has a positive and statistically significant effect on college attendance (\(\hat{\pi}_1 = 0.0098\), \(p < 0.001\)). The first-stage F-statistic is 390, which is far above the rule-of-thumb threshold of 10. Therefore, the instrument satisfies the relevance condition, meaning that \(Cov(scholarship, college) \neq 0\).

- **Estimates:** The IV estimate of the effect of college on income is approximately 21.081, compared to the OLS estimate with controls of 28.391. The lower IV estimate suggests that the OLS estimate may have been upward biased due to selection effects.

- **Should the IV estimate be interpreted as causal in this setting?** For the IV estimate to be interpreted causally, the instrument must satisfy the relevance and exogeneity assumptions. If scholarship is randomly assigned, the exogeneity assumption $Cov(scholarship,u)=0$ is plausible, while the first-stage results indicate that the instrument is relevant. Under these conditions, the IV estimate can be interpreted as causal.


# Question 3: Effect Heterogeneity

**(a) Which coefficient captures the heterogeneous effect, and how is it interpreted? **

We estimate the following interaction model:


$$\text{income}_i = \beta_0 + \beta_1 \text{college}_i + \beta_2 \text{white}_i + \beta_3 (\text{college}_i \times \text{white}_i) + u_i$$


The heterogeneous effect is captured by the interaction coefficient ($\beta_3$). It measures how the effect of college on income differs between white and non-white individuals. Following Wooldridge’s interpretation of interaction terms, the interaction term implies that the marginal effect of college on income no longer has to be constant across individuals. For non-white individuals ($\text{white} = 0$), the marginal effect of college is ($\beta_1$), while for white individuals ($\text{white} = 1$), the marginal effect is $\beta_1 + \beta_3$.

**(b) Estimate the equation from (a) in R.**

```{r interaction_model, include=TRUE}

interaction_model <- felm(
  income ~ 1 + college + white + college:white,
  data = df
)

summary(interaction_model, robust = TRUE)

```

**(c) Report the marginal effect of college on income separately.**

- If $\text{white} \mathrel{!=} 1$, the marginal effect of college on income is $\beta_1 = 38.072$. 
- If $\text{white} = 1$, the marginal effect is $\beta_1 + \beta_3 = 38.072 + 1.132 = 39.204$.

**(d) Is the difference in the college income premium across race groups statistically significant? What does this tell us about effect heterogeneity in the return to education?**

- No, the difference in the college income premium across race groups is not statistically significant at conventional significance levels (\( \alpha = 5\% \)), since the interaction coefficient $(\text{college}_i \times \text{white}_i)$ has a $(p)$-value of 0.216.

- Because both `college` and `white` are dummy variables, the interaction term measures the difference in the marginal effect of college on income between white and non-white individuals. The insignificant interaction effect suggests that we do not find strong evidence of effect heterogeneity in the return to education across race groups.


# Question 4: Probit and Logit Models

**(a) State one specific limitation of using OLS, when the dependent variable is binary.**

One limitation of using OLS with a binary dependent variable is that it can predict values below 0 or above 1, these fall outside the valid $[0,1]$ probability range and cannot be interpreted as meaningful probabilities.

**(b) Estimate a probit and a logit model in R using household income and race as covariates.**

```{r probit_logit_models, include=TRUE}

# Probit model
probit_model <- glm(
  college ~ 1 + hhincome + white,
  family = binomial(link = "probit"),
  data = df
)

summary(probit_model)

# Logit model
logit_model <- glm(
  college ~ 1 + hhincome + white,
  family = binomial(link = "logit"),
  data = df
)

summary(logit_model)

```


| Variable | Probit Model | Logit Model | Statistical Significance |
|---|---|---|---|
| `hhincome` | \(>0\) (0.010) | \(>0\) (0.017) | Significant in both models (\(p<0.001,\ \alpha=5\%\)) |
| `white` | \(<0\) (-0.105) | \(<0\) (-0.185) | Significant in both models (\(p<0.001,\ \alpha=5\%\)) |


- **Interpretation:** Both the probit and logit models agree on the direction and statistical significance of the effects. `hhincome` has a positive and statistically significant coefficient in both models ($p<0.001$), while `white` has a negative and statistically significant coefficient ($p<0.001$), conditional on household income. Since probit and logit are nonlinear models, the coefficients cannot be interpreted directly as marginal effects. While the direction of the effects is identified (e.g. `hhincome` is associated with a higher probability of attending college), the size (magnitude) of the effect depends on the values of the covariates and must be calculated separately.


# Question 5: Marginal Effects

**(a) Average Partial Effect (APE)**

The Average Partial Effect (APE) of a continuous variable \(x_j\) in a probit model is given by

$$
\frac{1}{n}
\sum_{i=1}^{n}
g\!\left(
\hat{\beta}_0
+
\hat{\beta}_1 x_{i1}
+
\cdots
+
\hat{\beta}_k x_{ik}
\right)
\hat{\beta}_j
$$

where \(g(\cdot)\) denotes the standard normal density function. The APE measures the average marginal effect of \(x_j\) on the probability \(P(Y=1 \mid X)\) across all observations.


**(b) Compute and interpret the APE of `hhincome` from the probit model estimated in 4(b).**


```{r probit, include=TRUE}

# Probit model
avg_slopes(probit_model, variables = "hhincome")

```

The estimated APE of `hhincome` is 0.00316 and is statistically significant (\(p < 0.001,\ \alpha = 5\%\)). This implies that a one-unit increase in household income increases the probability of attending college by approximately 0.316 percentage points on average.


**(c) Compute the predicted probability of attending college for an individual with `hhincome` == 80 and `white` == 1.**

```{r probit calc, include=TRUE}

predict(
  probit_model,
  newdata = data.frame(hhincome = 80, white = 1),
  type = "response"
)


```

The predicted probability of attending college for an individual with (`hhincome` = 80) and (`white` = 1) is approximately 74.0%. As expected in a probit model, the predicted probability lies between 0 and 1.

